---
title: "BST 270 Individual Project"
author: "Lauren Flynn"
date: "2023-01-17"
output: html_document
---

We will be examining the reproducibility of fivethirtyeight articles using the following example from a [fivethirtyeight article](https://fivethirtyeight.com/features/which-state-has-the-worst-drivers/) about "bad drivers", which utilizes information about vehicle collisions and car insurance to examine driver safety.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load libraries}
library(tidyverse)
library(RCurl)
```


The data is loaded from GitHub, where a chart displays the source information for each column. In the graphs that we will be reproducing below, the original source of the data is the National Highway Traffic Safety Administration (NHTSA). It is not clear based on the GitHub or the fivethirtyeight article how the data was wrangled to produce this dataset. Importantly, the NHTSA is a survey dataset, so it is not clear if the weights have been properly incorporated into the dataset. This dataset was accessed on Tuesday, January 17, 2023. It was last updated December 16, 2022. 


```{r load data}
gitData <- getURL("https://raw.githubusercontent.com/fivethirtyeight/data/master/bad-drivers/bad-drivers.csv")
carCrashes <- read.csv(text = gitData)
names(carCrashes)
```
The first figure we will reproduce is titled "Drivers Involved In Fatal Collisions While Speeding As a share of the number of fatal collisions per billion miles, 2009". 
```{r speeding collisions}
#add a column of counts of speeding related collisions
carCrashes <- carCrashes %>% mutate(fatalSpeedingAccidentsPerBillionMiles = Number.of.drivers.involved.in.fatal.collisions.per.billion.miles * Percentage.Of.Drivers.Involved.In.Fatal.Collisions.Who.Were.Speeding * 0.01)

#generate visualization
ggplot(data=carCrashes) +  
  geom_col(aes(x=State, y=Number.of.drivers.involved.in.fatal.collisions.per.billion.miles, fill = "Total Collisions Per Billion Miles"))  + 
  geom_col(aes(x=State, y= fatalSpeedingAccidentsPerBillionMiles, fill = "Speeding Collisions Per Billion Miles")) +
  scale_x_discrete(limits = rev) +
  coord_flip() + 
  scale_fill_manual(values=c("red", "pink")) + 
  labs(x="", y="", title = "Drivers Involved In Fatal Collisions While Speeding", subtitle = "As a share of the number of fatal collisions per billion miles, 2009", legend = "")  
```

The second figure we will reproduce is titled "Drivers Involved In Fatal Collisions While Alchol-Impaired As a share of the number of fatal collisions per billion miles, 2012". 
```{r alcohol collisions}
#add a column of counts of alcohol related collisions
carCrashes <- carCrashes %>% mutate(fatalAlcoholAccidentsPerBillionMiles = Number.of.drivers.involved.in.fatal.collisions.per.billion.miles * Percentage.Of.Drivers.Involved.In.Fatal.Collisions.Who.Were.Alcohol.Impaired * 0.01)


ggplot(data=carCrashes) +  
  geom_col(aes(x=State, y=Number.of.drivers.involved.in.fatal.collisions.per.billion.miles, fill = "Total Collisions Per Billion Miles"))  + 
  geom_col(aes(x=State, y= fatalAlcoholAccidentsPerBillionMiles, fill = "Alcohol Collisions Per Billion Miles")) +
  scale_x_discrete(limits = rev) +
  coord_flip() + 
  scale_fill_manual(values=c("deepskyblue3", "light blue")) + 
  labs(x="", y="", title = "Drivers Involved In Fatal Collisions While Alcohol-Impaired", subtitle = "As a share of the number of fatal collisions per billion miles, 2012")  
```



There is not a clear explanation on why different years were selected for the visualizations. 